---
title: "Human Activity Recognition Tool to Classify the Type of Weight Lifting Exercises" 
author: "Moh A"
date: "01/04/2017"
output: 
  html_document: 
    keep_md: yes
---

# Executive Summary

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, **the goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants to predict the manner in which they did the exercise**. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. The participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions: exactly according to the specification (Class A), while the other 4 classes (B, C, D and E) correspond to common mistakes.

Read more: http://groupware.les.inf.puc-rio.br/har#ixzz4d1GCDX2m

# Data PreProcessing

## Load Libraries
```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(caret)
```

## Download the Data

```{r, echo=TRUE, message=FALSE, warning=FALSE}
dataset_url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
dataset_filename <- "./data/pml-training.csv"

if (!file.exists("./data")){
    dir.create("./data")
}
if(!file.exists(dataset_filename)){
    download.file(dataset_url, destfile = dataset_filename, method = "curl")
}
```


## Read the Data

```{r, message=FALSE, warning=FALSE, include=FALSE}
raw_data <- read.table("./data/pml-training.csv", header = TRUE, sep = ",",  na.strings = c(NA, ""))

summary(raw_data)
mojority_NA_columns <- which(sapply(raw_data, function(x) sum(is.na(x)) >(nrow(raw_data)*0.9)))
cleaned_data <- raw_data[, -mojority_NA_columns]
cleaned_data <- cleaned_data[, -(1:7)]

set.seed(55555)
inTrain <- createDataPartition(y=cleaned_data$classe,
p=0.7, list=FALSE)
training <- cleaned_data[inTrain,]
testing <- cleaned_data[-inTrain,]

train_control <- trainControl(method="cv", number=10)
model_RF <- train(classe~., data = training, methods = "rf",trControl = train_control)

predict_RF <- predict(model_RF, testing)
confusionMatrix(testing$classe, predict_RF)


```


- how you built your model,
    - Different models tried
    - Comparison and evaluation of the different models.
- how you used cross validation, 
- what you think the expected out of sample error is,
- why you made the choices you did.

Reference

- Ugulino, W.; Cardador, D.; Vega, K.; Velloso, E.; Milidiu, R.; Fuks, H. Wearable Computing: Accelerometers' Data Classification of Body Postures and Movements. Proceedings of 21st Brazilian Symposium on Artificial Intelligence. Advances in Artificial Intelligence - SBIA 2012. In: Lecture Notes in Computer Science. , pp. 52-61. Curitiba, PR: Springer Berlin / Heidelberg, 2012. ISBN 978-3-642-34458-9. DOI: 10.1007/978-3-642-34459-6_6. 
