---
title: "Human Activity Recognition Tool to Classify the Type of Weight Lifting Exercises" 
author: "Moh A"
date: "01/04/2017"
output: 
  html_document: 
    keep_md: yes
---

# Executive Summary



## Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, **the goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants to predict the manner in which they did the exercise**. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. The participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions: exactly according to the specification (Class A), while the other 4 classes (B, C, D and E) correspond to common mistakes.

Read more: http://groupware.les.inf.puc-rio.br/har#ixzz4d1GCDX2m

# Exploratory Data Analysis

In the dataset given, there is 19622 observations and 160 variable.

The dependent variable, 'Classe', is a categorical variable corresponding to the five different ways of doing the weight lifting exercise: exactly according to the specification (Class A), throwing the elbows to the front (Class B), lifting the dumbbell only halfway (Class C), lowering the dumbbell only halfway (Class D) and throwing the hips to the front (Class E).

The rest of the variable corresponds to measurements from date, time, belt, arm, forearm and dumbbell sensors.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
classe_table <- table(raw_data$classe)
classe_percent <- (100*classe_table)/length(raw_data$classe)
kable(rbind(classe_table, classe_percent))
```

As seen in the table above, class 'A' (corresponding to correctly performing the exercise) is most frequent class and counts to roughly 28% among the five classes available. 

It is also worth noticing that frequency of class 'A' and hence the quality of performing the exercise varies significantly among the six participant, as evident in the barplot below.

```{r}
name_per_classe <-table(raw_data$classe, raw_data$user_name)

barplot(name_per_classe,
xlab = "Subject",
ylab = "Frequency",
main="Exercise Quality by Subject",
col = as.factor(rownames(name_per_classe)),
beside=TRUE
)

legend("topright" ,legend= rownames(name_per_classe), fill  = as.factor(rownames(name_per_classe)))
```

The aim of the classification model is to classify if exercise was completed correctly (class A) as well as classifying the type of error made ( class: B, C, D and E). Since the task given was to predict the class of 20 unknown observations. A model with an accuracy higher than 95% is needed. For example, the probability of a model with 95% accuracy to predict all cases (20/20) correctly is $p{20} = 0.95^20 = 0.36$. In contrast, a model with 99% accuracy will achieve $0.99^20 = 0.8$ probability in predicting the 20 cases correctly. 

# Data PreProcessing

## Load Libraries
```{r, message=FALSE, warning=FALSE, include=FALSE}
library(caret)
library(ggplot2)
library(knitr)
library(VIM)
```

## Download  and Read the Data

```{r, message=FALSE, warning=FALSE, include=FALSE}
dataset_url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
validation_url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
dataset_filename <- "./data/pml-training.csv"
validation_filename <- "./data/pml-testing.csv"

if (!file.exists("./data")){
    dir.create("./data")
}
if(!file.exists(dataset_filename)){
    download.file(dataset_url, destfile = dataset_filename, method = "curl")
}

if(!file.exists(validation_filename)){
    download.file(validation_url, destfile = validation_filename, method = "curl")
}

raw_data <- read.table("./data/pml-training.csv", header = TRUE, sep = ",",  na.strings = c(NA, ""))

validation_set <- read.table("./data/pml-testing.csv", header = TRUE, sep = ",",  na.strings = c(NA, ""))
```

The raw data file has 19622 observations and 160 variable.

## Clean the data

Firstly, looking at the plot below for the proportion of missing data among the dataset variables, it can be noticed that it has many variables with mostly missing and meaningless values (100 out of the 160 variables). Thus, their contribution to the predition of the 'classe' variable is insignificant and we have to omit these variables.

Secondly, it is wise to remove variables that do not contribute to the prediction of the 'classe' variable. Those incluse non-sensory measurements like: 'X'(row number), time and date related variables.

The rest of the variables have no missing values and are ready for further processing.


```{r, echo=FALSE, message=FALSE, warning=FALSE}
aggr(raw_data, prop=T, numbers=T)
mojority_NA_columns <- which(sapply(raw_data, function(x) sum(is.na(x)) >(nrow(raw_data)*0.9)))
cleaned_data <- raw_data[, -mojority_NA_columns]
cleaned_data <- cleaned_data[, -c(1,3,4,5,6)]
```


# Data Modeling

## Cross Validation & Out of Sample Error Estimation

K-fold cross validation is used to estimate the out of sample error, which is a reasonable balance between prediction accuracy and the computation time. The advantage of this method is that it matters less how the data gets divided. Every data point gets to be in a test set exactly once, and gets to be in a training set k-1 times. The variance of the resulting estimate is reduced as k is increased. The disadvantage of this method is that the training algorithm has to be rerun from scratch k times, which means it takes k times as much computation to make an evaluation. The results are then aggregated to create an overall estimate of the out of sample error. 

### Model 1: Linear Discriminant Analysis 

This model was selected since it is a relatively simple classification algorithm with fast training and prediction speed. However, it has a low average predictive accuracy and does not automatically learns feature interactions. Therefore, it could help in serving as a baseline for prediction accuracy.

The model scored an overall accuracy of 99.9%, with the highest sensitivity being 1 for classifying an exercise as class A when it is indeed A. The model also achieved high sensitivity values for other classes (B, C, D and E) as well. The confusion matrix illustrates that a classification model based on linear discriminant analysis have sufficient accuracy to expect perfect or near-perfect classification of our unknown validation cases.

### Model 2: Random Forest



```{r, message=FALSE, warning=FALSE, include=FALSE}
library(parallel)
library(doParallel)
library(mlbench)

set.seed(55555)
inTrain <- createDataPartition(y=cleaned_data$classe,
p=0.7, list=FALSE)
training <- cleaned_data[inTrain,]
testing <- cleaned_data[-inTrain,]

cluster <- makeCluster(detectCores() ) # convention to leave 1 core for OS
registerDoParallel(cluster)

intervalStart <- Sys.time()
lda_control <- trainControl(method = 'cv', number = 5, allowParallel = TRUE)
lda_model <- train(classe~., data = training, methods= 'lda', trControl = lda_control)
intervalEnd <- Sys.time()

paste("Linear Discriminant Analysis Model took: ", intervalEnd- intervalStart, attr(intervalEnd- intervalStart, "units"))

lda_test_predict <- predict(lda_model, testing)
confusionMatrix(lda_test_predict, testing$classe)

train_control <- trainControl(method="cv", number=10, allowParallel = TRUE)
model_RF <- train(classe~., data = training, methods = "rf",trControl = train_control)

stopCluster(cluster)
registerDoSEQ()

predict_RF <- predict(model_RF, testing)
confusionMatrix(testing$classe, predict_RF)


```


- how you built your model,
    - Different models tried
    - Comparison and evaluation of the different models.
- how you used cross validation, 
- what you think the expected out of sample error is,
- why you made the choices you did.

Reference

- Ugulino, W.; Cardador, D.; Vega, K.; Velloso, E.; Milidiu, R.; Fuks, H. Wearable Computing: Accelerometers' Data Classification of Body Postures and Movements. Proceedings of 21st Brazilian Symposium on Artificial Intelligence. Advances in Artificial Intelligence - SBIA 2012. In: Lecture Notes in Computer Science. , pp. 52-61. Curitiba, PR: Springer Berlin / Heidelberg, 2012. ISBN 978-3-642-34458-9. DOI: 10.1007/978-3-642-34459-6_6. 
